---
title: "PolyVAR Demo: Separating True vQTLs from mQTL Artefacts"
subtitle: "n = 250,000 — Three pipelines: Naive RINT · Wang 2019 (no-RINT) · PolyVAR"
author: "PolyVAR Development Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: { collapsed: false, smooth_scroll: true }
    toc_depth: 3
    self_contained: true
    fig_width: 14
    fig_height: 6
params:
  n:            250000
  seed:         2026
  M_mqtl:       30
  M_null:       19
  M_causal:     1
  beta_mqtl:    0.50
  beta_causal:  0.50
  delta_causal: 0.15
  maf:          0.25
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
suppressPackageStartupMessages({
  library(polyvar); library(ggplot2); library(data.table)
})
```

```{r constants}
set.seed(params$seed)
N       <- params$n
p       <- params$maf; q <- 1-p; sq <- sqrt(2*p*q)
BETA_M  <- params$beta_mqtl
BETA_C  <- params$beta_causal
DELTA   <- params$delta_causal
M_MQTL  <- params$M_mqtl
M_NULL  <- params$M_null
M_CAU   <- params$M_causal
M_TOTAL <- M_MQTL + M_NULL + M_CAU
CAUSAL_J <- M_MQTL + M_NULL + 1L

MCOLS <- c(
  "PolyVAR"          = "#1A5276",
  "PolyVAR-lite"     = "#5DADE2",
  "Levene (noRINT)"  = "#1E8449",
  "BF (noRINT)"      = "#52BE80",
  "Levene (naive)"   = "#C0392B",
  "BF (naive)"       = "#E67E22",
  "Bartlett (naive)" = "#8E44AD"
)
TCOLS <- c("mQTL only"="#E74C3C","Null"="#95A5A6","True vQTL"="#1A5276")
METHOD_LEVELS <- names(MCOLS)
TIER_MAP <- c(
  "PolyVAR"          = "PolyVAR",
  "PolyVAR-lite"     = "PolyVAR",
  "Levene (noRINT)"  = "Wang 2019 (no RINT)",
  "BF (noRINT)"      = "Wang 2019 (no RINT)",
  "Levene (naive)"   = "Naive RINT",
  "BF (naive)"       = "Naive RINT",
  "Bartlett (naive)" = "Naive RINT"
)
```

```{r simulate, cache=TRUE, cache.extra=params$seed}
phi_inv <- make_phi_inv(N)
lut     <- get_default_lut(N)

snp_info <- data.table(
  j       = seq_len(M_TOTAL),
  snp     = paste0("SNP", seq_len(M_TOTAL)),
  type    = c(rep("mQTL only",M_MQTL), rep("Null",M_NULL), rep("True vQTL",M_CAU)),
  beta    = c(rep(BETA_M,M_MQTL), rep(0,M_NULL), rep(BETA_C,M_CAU)),
  delta   = c(rep(0,M_MQTL+M_NULL), rep(DELTA,M_CAU)),
  is_vqtl = c(rep(FALSE,M_MQTL+M_NULL), rep(TRUE,M_CAU))
)

G <- matrix(rbinom(N*M_TOTAL, 2L, p), N, M_TOTAL)

run_snp <- function(j) {
  g_raw   <- G[,j]; g_star <- (g_raw-2*p)/sq
  sigma_i <- sqrt(1 + snp_info$delta[j]^2 * g_star^2)
  r <- snp_info$beta[j]*g_star + sigma_i*rnorm(N)
  r <- (r-mean(r))/sd(r)
  grp <- get_groups(as.numeric(g_raw))

  # Tier 1: Naive — RINT raw LOCO residual, no OLS
  er_naive <- apply_rint(r, phi_inv)
  lev_n  <- levene_rint_cpp(er_naive, grp$g0, grp$g1, grp$g2)
  bf_n   <- bf_rint_cpp(er_naive, grp$g0, grp$g1, grp$g2)
  bart_n <- bartlett_rint_cpp(er_naive, grp$g0, grp$g1, grp$g2)

  # Tier 2: Wang 2019 — OLS first, then Levene/BF without RINT
  e_ols <- ols_resid(r, g_star)$e
  lev_w <- levene_norint_cpp(e_ols, grp$g0, grp$g1, grp$g2)
  bf_w  <- bf_norint_cpp(e_ols, grp$g0, grp$g1, grp$g2)

  # Tier 3: PolyVAR
  pv  <- polyvar_full_cpp(r, g_raw, phi_inv)
  ls  <- polyvar_lite_setup_fast(r, 100L)
  pvl <- polyvar_lite_fast(r, g_raw, ls$breaks, ls$bin_q)

  data.table(
    j=j, snp=snp_info$snp[j],
    method=METHOD_LEVELS,
    T_stat=c(pv$T_C, pvl$T_C, lev_w$T_stat, bf_w$T_stat,
             lev_n$T_stat, bf_n$T_stat, bart_n$T_stat),
    pval  =c(pv$pval, pvl$pval, lev_w$pval, bf_w$pval,
             lev_n$pval, bf_n$pval, bart_n$pval)
  )
}

results <- rbindlist(lapply(seq_len(M_TOTAL), run_snp))
results <- merge(results, snp_info[,.(j,snp,type,delta,is_vqtl)], by=c("j","snp"))
results[, method  := factor(method,  levels=METHOD_LEVELS)]
results[, type    := factor(type,    levels=names(TCOLS))]
results[, tier    := factor(TIER_MAP[as.character(method)],
                            levels=c("Naive RINT","Wang 2019 (no RINT)","PolyVAR"))]
results[, pval_bh := p.adjust(pval,"BH"), by=method]
results[, sig_bh  := pval_bh < 0.05]
results[, neglogp := -log10(pmax(pval_bh,1e-300))]
results[, snp     := factor(snp, levels=paste0("SNP",seq_len(M_TOTAL)))]
```

---

## Scenario

```{r setup_tbl}
knitr::kable(data.table(
  Block=c(sprintf("Block 1: mQTL  (SNP1-SNP%d)",M_MQTL),
          sprintf("Block 2: Null  (SNP%d-SNP%d)",M_MQTL+1,M_MQTL+M_NULL),
          sprintf("Block 3: vQTL  (SNP%d)",M_TOTAL)),
  n=c(M_MQTL,M_NULL,M_CAU),
  beta=c(BETA_M,0,BETA_C), delta=c(0,0,DELTA),
  Truth=c("Null vQTL","Null vQTL","**True vQTL**"),
  Expected=c(
    "FP for naive RINT; partial fix by Wang 2019; zero FP for PolyVAR",
    "True negatives for all methods",
    "TP — all methods detect; only PolyVAR has clean FP profile in Block 1")),
  align="llcccl")
```

**Three pipelines:**

| Tier | Pipeline | Contamination NCP |
|---|---|---|
| **Naive RINT** | RINT on raw LOCO residual → Levene / BF / Bartlett | ∝ n·β⁴ — largest |
| **Wang 2019** | OLS first → Levene / BF without RINT | Reduced but not zero |
| **PolyVAR** | OLS → RINT → score-space FWL projection | **= 0 exactly** |

---

## Results

### Manhattan plot

```{r manhattan, fig.height=10, fig.width=14}
sig_line <- -log10(0.05)
ymax <- max(results$neglogp, na.rm=TRUE)*1.12

ggplot(results, aes(x=snp, y=neglogp, colour=type, shape=is_vqtl, size=is_vqtl)) +
  annotate("rect",xmin=0.5,xmax=M_MQTL+0.5,ymin=0,ymax=ymax,fill="#FDEDEC",alpha=0.30)+
  annotate("rect",xmin=M_MQTL+M_NULL+0.5,xmax=M_TOTAL+0.5,
           ymin=0,ymax=ymax,fill="#EBF5FB",alpha=0.40)+
  geom_hline(yintercept=sig_line,linetype="dashed",colour="grey20",linewidth=0.65)+
  annotate("text",x=1.2,y=sig_line+0.12,
           label="BH FDR=5%",hjust=0,size=2.9,colour="grey25")+
  annotate("label",x=M_MQTL/2,y=ymax*0.93,
           label=sprintf("Block 1\nmQTL\nbeta=%.2f",BETA_M),
           size=2.7,colour="#C0392B",fontface="bold",fill="white",label.size=0)+
  annotate("label",x=M_MQTL+M_NULL/2,y=ymax*0.93,
           label="Block 2\nNull",
           size=2.7,colour="#7F8C8D",fill="white",label.size=0)+
  annotate("label",x=M_TOTAL-0.3,y=ymax*0.93,
           label=sprintf("Block 3\nTrue vQTL\ndelta=%.2f",DELTA),
           size=2.7,colour="#1A5276",fontface="bold",fill="white",label.size=0,hjust=1)+
  geom_point(alpha=0.88)+
  scale_colour_manual(values=TCOLS,name="SNP type")+
  scale_shape_manual(values=c("FALSE"=16,"TRUE"=18),guide="none")+
  scale_size_manual(values=c("FALSE"=1.8,"TRUE"=4.5),guide="none")+
  scale_x_discrete(breaks=paste0("SNP",c(1,10,M_MQTL,M_MQTL+1,
                                          M_MQTL+M_NULL,M_TOTAL)))+
  scale_y_continuous(limits=c(0,ymax),expand=c(0,0))+
  facet_wrap(~method,nrow=3,ncol=3)+
  labs(title=expression("Manhattan: BH-adjusted "*-log[10](p)*
                        " — all 7 methods, three pipeline tiers"),
       subtitle=sprintf("n=%s  |  %d mQTL SNPs (Block 1, beta=%.2f)  |  true vQTL=SNP%d (diamond)  |  BH FDR=5%%",
                        format(N,big.mark=","),M_MQTL,BETA_M,M_TOTAL),
       x=NULL,y=expression(-log[10](p[BH])))+
  theme_bw(base_size=10)+
  theme(strip.background=element_rect(fill="#1B2631",colour=NA),
        strip.text=element_text(colour="white",face="bold",size=10),
        axis.text.x=element_text(angle=45,hjust=1,size=7),
        panel.grid.minor=element_blank(),panel.grid.major.x=element_blank(),
        legend.position="bottom",panel.spacing=unit(0.5,"lines"),
        plot.title=element_text(face="bold",size=13),
        plot.subtitle=element_text(size=8.5,colour="grey35"))
```

---

### T statistics across all SNPs

```{r tstat, fig.height=10, fig.width=14}
ymax_t <- max(results$T_stat,na.rm=TRUE)*1.05

ggplot(results,aes(x=snp,y=T_stat,colour=type,shape=is_vqtl,size=is_vqtl))+
  annotate("rect",xmin=0.5,xmax=M_MQTL+0.5,ymin=0,ymax=ymax_t,
           fill="#FDEDEC",alpha=0.30)+
  annotate("rect",xmin=M_MQTL+M_NULL+0.5,xmax=M_TOTAL+0.5,ymin=0,ymax=ymax_t,
           fill="#EBF5FB",alpha=0.40)+
  geom_hline(yintercept=2,linetype="dotted",colour="grey50",linewidth=0.6)+
  geom_point(alpha=0.88)+
  scale_colour_manual(values=TCOLS,name=NULL)+
  scale_shape_manual(values=c("FALSE"=16,"TRUE"=18),guide="none")+
  scale_size_manual(values=c("FALSE"=1.8,"TRUE"=4.5),guide="none")+
  scale_x_discrete(breaks=paste0("SNP",c(1,10,M_MQTL,M_MQTL+1,
                                          M_MQTL+M_NULL,M_TOTAL)))+
  scale_y_continuous(limits=c(0,ymax_t),expand=c(0,0))+
  facet_wrap(~method,nrow=3,ncol=3,scales="free_y")+
  labs(title="Raw test statistics T across all SNPs — three pipeline tiers",
       subtitle="Dotted = null expectation (2).  Pink = mQTL block.  Blue = true vQTL.  Y-axes free per panel.",
       x=NULL,y="Test statistic T")+
  theme_bw(base_size=10)+
  theme(strip.background=element_rect(fill="#1B2631",colour=NA),
        strip.text=element_text(colour="white",face="bold",size=10),
        axis.text.x=element_text(angle=45,hjust=1,size=7),
        panel.grid.minor=element_blank(),panel.grid.major.x=element_blank(),
        legend.position="bottom",panel.spacing=unit(0.5,"lines"),
        plot.title=element_text(face="bold",size=12),
        plot.subtitle=element_text(size=8.5,colour="grey35"))
```

---

### mQTL block vs true vQTL: distinguishability

```{r sidebyside, fig.height=5.5, fig.width=14}
cmp <- rbind(
  results[type=="mQTL only",
          .(method,tier,T_stat,grp="Block 1: mQTL (delta=0)\nNULL vQTL")],
  results[type=="True vQTL",
          .(method,tier,T_stat,grp=sprintf("Block 3: True vQTL (delta=%.2f)",DELTA))]
)
cmp[, grp := factor(grp,levels=c("Block 1: mQTL (delta=0)\nNULL vQTL",
                                   sprintf("Block 3: True vQTL (delta=%.2f)",DELTA)))]

ggplot(cmp,aes(x=grp,y=T_stat,fill=grp))+
  geom_boxplot(data=cmp[grp=="Block 1: mQTL (delta=0)\nNULL vQTL"],
               width=0.45,alpha=0.8,outlier.size=1.2)+
  geom_point(data=cmp[grp==sprintf("Block 3: True vQTL (delta=%.2f)",DELTA)],
             shape=18,size=6,alpha=1)+
  geom_hline(yintercept=2,linetype="dotted",colour="grey55",linewidth=0.7)+
  scale_fill_manual(
    values=setNames(c("#E74C3C","#1A5276"),
                    c("Block 1: mQTL (delta=0)\nNULL vQTL",
                      sprintf("Block 3: True vQTL (delta=%.2f)",DELTA))),
    guide="none")+
  facet_wrap(~method,nrow=2,ncol=4)+
  labs(title="mQTL block (boxplot) vs true vQTL (diamond) by method",
       subtitle=paste0(
         "Naive RINT: box and diamond at same height — indistinguishable.\n",
         "Wang 2019 (noRINT): box lower than naive, but still above null.\n",
         "PolyVAR: box at null (~2), diamond clearly isolated."),
       x=NULL,y="Test statistic T")+
  theme_bw(base_size=10.5)+
  theme(strip.background=element_rect(fill="#1B2631",colour=NA),
        strip.text=element_text(colour="white",face="bold",size=10.5),
        panel.grid.major.x=element_blank(),panel.grid.minor=element_blank(),
        panel.spacing=unit(0.5,"lines"),
        plot.title=element_text(face="bold",size=12),
        plot.subtitle=element_text(size=9,colour="grey35"),
        axis.text.x=element_text(size=9))
```

---

### Block 1 FPR and mean T by method

```{r tier_summary}
tier_sum <- results[type=="mQTL only", .(
  median_T = round(median(T_stat,na.rm=TRUE),1),
  mean_T   = round(mean(T_stat,  na.rm=TRUE),1),
  pct_sig  = round(mean(sig_bh,  na.rm=TRUE)*100,1)
), by=.(tier,method)]
setorder(tier_sum, tier, method)
knitr::kable(tier_sum,
  col.names=c("Tier","Method","Median T","Mean T","FPR % (BH 5%)"),
  align=c("l","l","c","c","c"),
  caption=paste0("Block 1 (delta=0, mQTL only) summary. Null expectation: T=2, FPR=5%."))
```

---

### Confusion matrix

```{r cm}
cm <- results[, .(
  TP    = sum(is_vqtl  &  sig_bh),
  FP    = sum(!is_vqtl &  sig_bh),
  FN    = sum(is_vqtl  & !sig_bh),
  TN    = sum(!is_vqtl & !sig_bh),
  FDR   = round(sum(!is_vqtl&sig_bh)/max(sum(sig_bh),1)*100,0),
  Power = round(sum(is_vqtl&sig_bh)/M_CAU*100,0)
), by=.(tier,method)]
cm[, Verdict := fcase(
  FP==0 & TP==M_CAU, "Correct",
  FP==0,             "Missed",
  TP==M_CAU,         "TP + FP",
  default,           "FP only")]
knitr::kable(cm,
  col.names=c("Tier","Method","TP","FP","FN","TN","FDR%","Power%","Verdict"),
  align=c("l","l","c","c","c","c","c","c","l"),
  caption=sprintf("Confusion matrix at BH FDR=5%%. True vQTL=SNP%d. Null SNPs=1-%d.",
                  M_TOTAL, M_TOTAL-1L))
```

---

### Smoking gun: residual density by genotype group

```{r mechanism, fig.height=4.5, fig.width=13}
j_demo <- 1L
g0 <- G[,j_demo]; gs0 <- (g0-2*p)/sq
r0 <- BETA_M*gs0 + rnorm(N); r0 <- (r0-mean(r0))/sd(r0)
bh0 <- sum(gs0*r0)/sum(gs0^2); e0 <- r0-bh0*gs0
er_n <- apply_rint(r0, phi_inv)
er_o <- apply_rint(e0, phi_inv)
glab <- c("AA (g=0)","Aa (g=1)","aa (g=2)")[g0+1L]

dt <- rbind(
  data.table(val=er_n, group=glab, pipe="Naive: RINT(r)"),
  data.table(val=e0,   group=glab, pipe="Wang 2019: OLS residual\n(no RINT)"),
  data.table(val=er_o, group=glab, pipe="PolyVAR: RINT(OLS residual)")
)
dt[, pipe := factor(pipe,levels=c("Naive: RINT(r)",
                                   "Wang 2019: OLS residual\n(no RINT)",
                                   "PolyVAR: RINT(OLS residual)"))]
vsum <- dt[, .(v=round(var(val),4)), by=.(pipe,group)]
y_pos <- rep(c(0.55,0.43,0.31),3)

ggplot(dt[, .SD[sample(.N,min(.N,3000))], by=.(pipe,group)],
       aes(x=val,fill=group,colour=group))+
  geom_density(alpha=0.28,linewidth=0.85)+
  geom_text(data=vsum,aes(x=Inf,y=y_pos,
    label=sprintf("%s: var=%.4f",sub(" .*","",group),v)),
    inherit.aes=FALSE,hjust=1.05,size=2.8,colour="grey20")+
  scale_fill_manual(values=c("AA (g=0)"="#2980B9","Aa (g=1)"="#27AE60",
                              "aa (g=2)"="#C0392B"),name="Genotype")+
  scale_colour_manual(values=c("AA (g=0)"="#2980B9","Aa (g=1)"="#27AE60",
                                "aa (g=2)"="#C0392B"),guide="none")+
  facet_wrap(~pipe,nrow=1)+
  labs(title=sprintf("Group-conditional residual density (SNP1, beta=%.2f, delta=0)",BETA_M),
       subtitle=paste0(
         "Left: naive RINT — group variances differ, Levene fires (FALSE POSITIVE).\n",
         "Middle: Wang 2019 OLS only — variances partially equalised, residual contamination remains.\n",
         "Right: PolyVAR RINT(OLS) — group variances equal, contamination = 0."),
       x="Residual value",y="Density")+
  theme_bw(base_size=11)+
  theme(strip.background=element_rect(fill="#1B2631",colour=NA),
        strip.text=element_text(colour="white",face="bold",size=11),
        panel.grid.minor=element_blank(),legend.position="bottom",
        plot.title=element_text(face="bold",size=12),
        plot.subtitle=element_text(size=9,colour="grey35"))
```

---

## Interpretation

**Naive RINT:** RINT applied to raw LOCO residual $r_i = \beta g^*_i + \varepsilon_i$ creates spurious group-conditional variance differences. Every Block 1 mQTL is called a vQTL.

**Wang et al. 2019 (no RINT):** OLS removes $\beta g^*$ before testing. This substantially reduces contamination because the residuals are approximately iid. However, because OLS is a linear operation and the residual contamination is a nonlinear function of estimation error, the Frisch-Waugh-Lovell theorem does not apply and some contamination NCP remains. At $n=250{,}000$ with $\beta=0.50$ this residual NCP still exceeds the BH threshold, giving intermediate FPR between naive and PolyVAR.

**PolyVAR:** OLS + RINT + score-space FWL projection. The projection $U_k^c = U_k - \frac{\widehat{\text{Cov}}(U_k,U_\mu)}{\widehat{\text{Var}}(U_\mu)} U_\mu$ is applied in score space where FWL holds exactly, setting the contamination NCP to zero for all $\beta$, $n$, and MAF.

---

## Session info

```{r session, echo=TRUE}
sessionInfo()
```
